Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer

### 2018-05-24 submission/manuscript.Rmd
final updates to first draft

# Load in data for select genes in SCFA pathways
picrust_select <- read_csv("../data/process/specific_scfa_kruskal_picrust_summary.csv")
opf_select <- read_csv("../data/process/select_scfa_opf_kruskal_summary.csv")

@@

**Background.** Certain bacterial community members have been shown to increase colon tumorigenesis while other taxa have been associated with normal colons. The some of the taxa associated with normal colons can use fiber to produce short-chain fatty acids which have been shown to inhibit tumor growth in model systems. However, the data supporting the importance of SCFAs in humans is less certain. Here we test the hypothesis that short-chain fatty acid concentration decreases in individuals with colorectal tumors.

@@

**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance based on individuals having normal colons, adenomas, or carcinomas. First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes inolved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2]. This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A]. Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and performed metagenomic sequencing to confirm these results. Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3]. This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 3B]. These observations provide evidence that the gene content does not change due to tumors. Although these results suggest that SCFA concentrations do not change due to tumor, there may be some errors present within our results that may account for the lack of associations.

>>>>

# Load in data for select genes in SCFA pathways
picrust_select <- read_csv("../data/process/tables/specific_scfa_kruskal_picrust_summary.csv")
picrust_select_treat <- read_csv("../data/process/tables/specific_scfa_wilcox_picrust_treatment_summary.csv")
opf_select <- read_csv("../data/process/tables/select_scfa_opf_kruskal_summary.csv")

@@

**Background.** Certain bacterial community members have been shown to increase colon tumorigenesis while other taxa have been associated with normal colons. The some of the taxa associated with normal colons can use fiber to produce short-chain fatty acids which have been shown to inhibit tumor growth in model systems. However, the data supporting the importance of SCFAs in humans is less certain. Here we test the hypothesis that short-chain fatty acid concentration changes in individuals with colorectal tumors.

@@

**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance based on individuals having normal colons, adenomas, or carcinomas. First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes inolved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2]. This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A]. Using a Wilcoxson rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3]. Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and performed metagenomic sequencing to confirm these results. Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4]. This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 3B]. These observations provide evidence that the gene content does not change due to tumors. Although these results suggest that SCFA concentrations do not change due to tumor, there may be some errors present within our results that may account for the lack of associations.

### 2018-05-24 submission/manuscript.tex
final updates to first draft

\textbf{Background.} Certain bacterial community members have been shown
to increase colon tumorigenesis while other taxa have been associated
with normal colons. The some of the taxa associated with normal colons
can use fiber to produce short-chain fatty acids which have been shown
to inhibit tumor growth in model systems. However, the data supporting
the importance of SCFAs in humans is less certain. Here we test the
hypothesis that short-chain fatty acid concentration decreases in
individuals with colorectal tumors.

@@

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance based on individuals having normal colons,
adenomas, or carcinomas. First, using imputed gene relative abundance
based on 16S rRNA gene sequencing we found no difference in any of the
genes inolved with acetate, butyrate, or propionate synthesis (P-value
\textgreater{} 0.90) {[}Table S2{]}. This similarity between groups is
highlighted by visualizing genes important in butyrate synthesis
{[}Figure 3A{]}. Next, we took a subset of these 490 fecal samples
(n=85) and performed metagenomic sequencing to confirm these results.
Like the imputed gene results, metagenomic analysis found that there was
no difference in any of the genes involved in SCFA synthesis between
individuals with normal colons (n=29) and patients with adenoma (n=28)
or carcinoma (n=28) (P-value \textgreater{} 0.70) {[}Table S3{]}. This
lack of difference is highlighted when we visualize the results for
butyrate kinase {[}Figure 3B{]}. These observations provide evidence
that the gene content does not change due to tumors. Although these
results suggest that SCFA concentrations do not change due to tumor,
there may be some errors present within our results that may account for
the lack of associations.

>>>>

\textbf{Background.} Certain bacterial community members have been shown
to increase colon tumorigenesis while other taxa have been associated
with normal colons. The some of the taxa associated with normal colons
can use fiber to produce short-chain fatty acids which have been shown
to inhibit tumor growth in model systems. However, the data supporting
the importance of SCFAs in humans is less certain. Here we test the
hypothesis that short-chain fatty acid concentration changes in
individuals with colorectal tumors.

@@

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance based on individuals having normal colons,
adenomas, or carcinomas. First, using imputed gene relative abundance
based on 16S rRNA gene sequencing we found no difference in any of the
genes inolved with acetate, butyrate, or propionate synthesis (P-value
\textgreater{} 0.90) {[}Table S2{]}. This similarity between groups is
highlighted by visualizing genes important in butyrate synthesis
{[}Figure 3A{]}. Using a Wilcoxson rank-sum test, there also was no
difference in imputed gene relative abundance between pre- and
post-treatment samples for any genes involved with SCFA synthesis
(P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a subset of
these 490 fecal samples (n=85) and performed metagenomic sequencing to
confirm these results. Like the imputed gene results, metagenomic
analysis found that there was no difference in any of the genes involved
in SCFA synthesis between individuals with normal colons (n=29) and
patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
0.70) {[}Table S4{]}. This lack of difference is highlighted when we
visualize the results for butyrate kinase {[}Figure 3B{]}. These
observations provide evidence that the gene content does not change due
to tumors. Although these results suggest that SCFA concentrations do
not change due to tumor, there may be some errors present within our
results that may account for the lack of associations.

### 2018-05-24 submission/supplemental.Rmd
final updates to first draft

**Table S3: Breakdown of Kruskal-Wallis tests using counts per million from metagenomic sequencing of select genes in SCFA pathways.**
```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
table_s3 <- read_csv("../data/process/select_scfa_opf_kruskal_summary.csv") %>% 
  select(kegg_id, statistic, p.value, bh) %>% 
  mutate_at(vars(statistic:bh), function(x) round(x, digits = 3))

colnames(table_s3) <-  c("KEGG IDs", "Rank Sum Statistic", "P-value", "BH")

@@

>>>>

**Table S3: Breakdown of Wilcoxson rank-sum tests for pre- versus post-treatment of tumor using PICRUSt imputed relative abundance of select genes in SCFA pathways.**
```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}

table_s3 <- read_csv("../data/process/tables/specific_scfa_wilcox_picrust_treatment_summary.csv") %>% 
  select(kegg_ortholog, group, statistic, p.value, bh) %>% 
   mutate_at(vars(p.value:bh), function(x) round(x, digits = 3))

colnames(table_s3) <-  c("KEGG IDs", "Disease Group", "Test Statistic", "P-value", "BH")

@@

**Table S4: Breakdown of Kruskal-Wallis tests using counts per million from metagenomic sequencing of select genes in SCFA pathways.**
```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
table_s4 <- read_csv("../data/process/select_scfa_opf_kruskal_summary.csv") %>% 
  select(kegg_id, statistic, p.value, bh) %>% 
  mutate_at(vars(statistic:bh), function(x) round(x, digits = 3))

colnames(table_s4) <-  c("KEGG IDs", "Rank Sum Statistic", "P-value", "BH")

kable(table_s4, format = "markdown", align = 'c', padding = 2)
```
